Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2021-09-01 write-up/covid-writeup.tex
Updated SEIR one last time

\begin{abstract}
  Coronavirus case-count data has influenced government policies and drives most epidemiological forecasts. Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic. While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the COVID-19 pandemic; neither can be fully addressed by increased testing capacity. In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number. We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.  To address this, a procedure is presented that combines case-count data and random samples over time to estimate selection propensities based on key covariate information. We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.  This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, a statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.  We end with a series of recommendations based on the proposed methodology.
\end{abstract}

@@

 After demonstrating the limitations of case-count analysis when compared to random sampling, we then assess whether their is potential for combining the \emph{nonprobability samples} with \emph{probability samples} to improve point prevalence estimation. For any probability sampling design, the Horvitz-Thompson estimator~\citep{HT1952} incorporates design information via inverse-probability weights (IPW).  For nonprobability samples, the IPW estimator requires modelling the propensity scores.  Its use in the survey context is also referred to as quasi-randomization~\citep{Elliott2017}. \cite{Valliant2011} consider a weighted logistic regression procedure using the pooled probability and nonprobability samples.  \cite{Chen2019} consider a pseudo-log-likelihood approach that uses the random samples as a proxy for a term in the log-likelihood.  Here, we extend this approach to account for measurement-error as well as observing random samples at multiple times. We then provide an extension of the statistical error decomposition and discuss the trade-offs inherent in such a weighting approach.

 One core component of coronavirus research is epidemiological compartmental modelling of case-count and death data. These models can be used to answer a variety of research questions including case-count forecasting, estimation of the effective reproduction number, and estimation of quarantine and other health policies on infectious disease dynamics.  The basic approach is a deterministic compartmental model called the susceptible-infectious-recovered (SIR) model.  A probabilistic extension was proposed by~\cite{Osthus2017} to model one-dimensional time series of infected proportions. \cite{Song2020} extended this approach to incorporate interventions and assess interventions on COVID-19 epidemic in China. \cite{Hao2020} builds an extended SEIR model (SAPHIRE) to account for various presymptomatic infectiousness, time-varying ascertainment rates, transmission rates and population movements.

Given a probability sampling design, individual predictions can be leveraged to improve estimation via model-assisted approaches~\citep{Breidt2017}.  For nonprobability samples, \cite{Chen2019} derive a doubly robust approach that uses outcome predictions given covariates on the nonprobability and probability samples.  Here, we combine the state-space SEIR approach of \cite{Song2020} but instead, as in \cite{Johndrow2020}, focus on COVID-19 confirmed death count data.  We generate epidemiological forecasts for point prevalence of each population strata.  We then demonstrate how to combine these forecasts with the IPW approach to construct doubly-robust estimates of point prevalence.  A derived statistical error decomposition guides this discussion.

@@

\begin{remark}[Reporting Delays]
Figure~\ref{fig:in-data} shows clear reduced testing and case count reporting on weekends compared to weekdays. Moreover, COVID-19 tests are reported on the day they were administered, while case counts are reported based on the date the positive test was reported to and confirmed by the Indiana Department of Health system.  To try and address both issues, analyses are conducted on the weekly level in this manuscipt.
 \end{remark}

@@

 Between April 25--29, 2020, Indiana conducted statewide random molecular testing of persons ages $\geq 12$ years to assess prevalence of active infection to SARS-CoV-2~\citep{Yiannoutsos2021}. A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata. 15,495 participants were contacted resulting in a final sample size of 3,658. Demographic data was collected (e.g., summary statistics on age, sex, and race) as well as data on whether they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2. During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.
 % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.

@@

 We start with some simple notation.  Let $N$ denote the population size.  At a given time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$. Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest. For simplicity, we start by ignoring the dynamic nature of the outbreak and recoverability of individuals. We assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$. We also let $I_j \in \{0,1\}$ be an indicator that the individual was tested ($I = 1$) or not ($I=0$).

 To start, we assume the overall number of active COVID-19 cases and/or active infection rate (AIR) are of primary interest. That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$. Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.  Then a natural candidate for AIR is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall active cases is $N \times \bar y$.
 Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.  Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.

@@

 Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and log odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}. Assuming no measurement error, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., if COVID-19 positive individuals were more likely to receive test then this implies upward bias in prevalence estimates. Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply magnifies the error. When tests are imperfect and selection bias exists, this simple relationship no longer holds.

@@

 \begin{figure}
 \centering
 \begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/sir_ratio.png}
  \caption{Ratio estimator}
  \label{fig:ratio-bias}
 \end{subfigure}%
 \begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/sir_rt.png}
  \caption{Effective reproductive rate estimator}
  \label{fig:r0-bias}
 \end{subfigure}
 \caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.  Here we assume $f = 0.02$, $FP = 0.024$, $FN = 0.13$, and use a range of relative sampling fractions $M = f_1/f_0$.}
 \label{fig:rates}
 \end{figure}

@@

 % While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
 Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.  First, we can write the difference in the rates as $R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)$ where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.  Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.  Here, these errors can mingle in interesting ways.  Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.  Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model (A=black and B=red). Of interest is the difference between the effective reproduction rate for country A and B, $R_{tA} - R_{tB}$. Figure~\ref{fig:eff} shows how biases in estimation of each rate interact in complex ways.  At first, the difference is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the magnitude of over-estimation increases as country A sees declining case-count while country B sees rapidly increasing case-count; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees declining case-count; finally, the gap disappears.  We do not claim this will always be the case; however, this demonstrates how observed information can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).

@@

 To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.  For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now. Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$. Maximum likelihood estimation follows by maximizing
 \begin{equation}
 \label{eq:propensity}
 \sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
 \end{equation}
 The first sum only involves individuals observed in the nonprobability sample.
 The second sum is over the entire population.  Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.  Typically, this is not possible.  Here we present a method that uses auxiliary information obtained from probability samples.

@@

 As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.  Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.  Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.   Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer data size.

@@

 If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.  Unfortunately for COVID-19, large probabilistic samples are not available at every time within a given region.  Here we consider a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
 $$
 \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
 $$
 where $K_h$ is a kernel function with tunable parameter $h$. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest} using case-count data observed on day~$t$.

 % Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
 % $$
 % \hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \tilde d_{k,t^\prime}}
 % $$
 % where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

 %\subsubsection{Uncertain sensitivity and specificity}
 %
 %For prevalence there is some
 %
 %For
 %
 %For seroprevalence, \cite{Bendavid2020} report
 %
 %
 %
 %, which leads to a range for $M$ of $(2.21, 2.38)$.  Therefore, the impact on bias of $\hat R_t$ appears moderate.  If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.  This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
 %
 %
 \subsubsection{Asymptotics}
 \label{section:asymptotics}

 Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.  Each finite population has an associated non-probability sample at time $t$ of size~$\tilde n$ and $L_\nu$ probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$. Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent as the number of probability samples grows and calculates the estimator's variance.  In Lemma~\ref{lemma:ipw}, sensitivity and specificity are not assumed known.  Instead, we suppose that these are estimated using a pseudo-likelihood on two random samples from the population whose size (denoted~$n_{FP}$ and $n_{FN}$ respectively) is chosen so the resulting estimated sensitivity and specificity have desired levels of uncertainty.  See Appendix~\ref{app:asympderivations} for additional details.

 \begin{lemma}[Variance of IPW estimator] \normalfont
 \label{lemma:ipw}
 % \textcolor{red}{Must check.}
 The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ are solutions to the following set of estimating equations
 $$
 \Phi_n (\eta_t) =
 \left (
 \begin{array}{c}
 \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
 \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
 \frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
 \frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
 \end{array}
 \right ) = {\bf 0}
 $$
 where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{t,t_l}$ and $V_t^{(IPW)}$ is the first diagonal element of $E [\Phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\Phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
 \end{lemma}

@@

 Selection bias is addressed by analyzing COVID-19 death data rather than case counts.  Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$.  Then
 $$
 D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r \text{IFR}_{k} \cdot I_{k,t} \theta_{(r-t)} \right)
 $$
 where $\text{IFR}_{k}$ is the infection fatality rate for the $k$th strata and $\theta_{j}$ is a discrete-time distribution for time from infection to death.  This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on a stratification variable, which is important as IFR depends heavily on age.  Specifically, ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85''~\cite{Levin2020}. For conciseness, parameter choices and additional details on inference are discussed in Appendix~\ref{app:modelbased}.

@@

 In the first approach, we estimate two logistic regressions using weighted pseudo-likelihoods with the Facebook survey symptom data.  The first estimates the probability of contact with a COVID-19 positive individual given demographic and COVID-19 test status (positive or negative COVID-19 test). Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of contact with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.  The second estimates the probability of fever given demographic, COVID-19 test status, \emph{and} whether the individual has had contact with a COVID-19 positive individual.  Figure~\ref{fig:symptomlik1} and~\ref{fig:symptomlik2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.  We see that likelihood of fever varies greatly based on whether the individual also reported contact with a COVID-19 positive individual. Using these two models, mean imputation is used to calculate the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status. Inverse probability weights are then computed based on the resulting dataset (termed IPW1).

>>>>

\begin{abstract}
  Coronavirus case-count data has influenced government policies and drives most epidemiological forecasts. Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic. While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the COVID-19 pandemic; neither can be fully addressed by increased testing capacity. In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number. We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.  To address this, a procedure is presented that combines case-count data and random samples over time to estimate selection propensities based on key covariate information. We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.  This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with demographic information, a statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.  We end with a series of recommendations based on the proposed methodology.
\end{abstract}

@@

 After demonstrating the limitations of case-count analysis when compared to random sampling, we then assess whether there is potential for combining the \emph{nonprobability samples} with \emph{probability samples} to improve point prevalence estimation. For any probability sampling design, the Horvitz-Thompson estimator~\citep{HT1952} incorporates design information via inverse-probability weights (IPW).  For nonprobability samples, the IPW estimator requires modelling the propensity scores.  Its use in the survey context is also referred to as quasi-randomization~\citep{Elliott2017}. \cite{Valliant2011} consider a weighted logistic regression procedure using the pooled probability and nonprobability samples.  \cite{Chen2019} consider a pseudo-log-likelihood approach that uses the random samples as a proxy for a term in the log-likelihood.  Here, we extend this approach to account for measurement-error as well as observing random samples at multiple times. We then provide an extension of the statistical error decomposition and discuss the trade-offs inherent in such a weighting approach.

 One core component of coronavirus research is epidemiological compartmental modelling of case-count and death data. These models can be used to answer a variety of research questions including case-count forecasting, estimation of the effective reproduction number, and estimation of quarantine and other health policies on infectious disease dynamics.  The basic approach is a deterministic compartmental model called the susceptible-infectious-recovered (SIR) model.  A probabilistic extension was proposed by~\cite{Osthus2017} to model one-dimensional time series of infected proportions. \cite{Song2020} extended this approach to incorporate interventions and assess interventions on COVID-19 epidemic in China. \cite{Hao2020} extends this work further to account for various presymptomatic infectiousness, time-varying ascertainment rates, transmission rates and population movements.

Given a probability sampling design, individual predictions can be leveraged to improve estimation via model-assisted approaches~\citep{Breidt2017}.  For nonprobability samples, \cite{Chen2019} derive a doubly robust approach that uses outcome predictions given covariates on the nonprobability and probability samples.  Here, we combine the compartmental model of \cite{Song2020} but instead, as in \cite{Johndrow2020}, focus on COVID-19 confirmed death count data.  We generate epidemiological forecasts for point prevalence of each population strata.  We then demonstrate how to combine these forecasts with the IPW approach to construct doubly-robust estimates of point prevalence.  A derived statistical error decomposition guides this discussion.

@@

\begin{remark}[Reporting Delays]
Figure~\ref{fig:in-data} shows clear reduced testing and case count reporting on weekends compared to weekdays. Moreover, COVID-19 tests are reported on the day they were administered, while case counts are reported based on the date the positive test was reported to and confirmed by the Indiana Department of Health system.  To minimize the impact of reporting on analysis, testing and case-count data are aggregated at the weekly level for analysis.
 \end{remark}

@@

 Between April 25--29, 2020, Indiana conducted statewide random molecular testing of persons ages $\geq 12$ years to assess prevalence of active infection to SARS-CoV-2~\citep{Yiannoutsos2021}. A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata. 15,495 participants were contacted resulting in a final sample size of 3,658. Demographic data was collected (e.g., summary statistics on age, sex, and race) as well as data on whether they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2. During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.
 Table~\ref{tab:indiana} summarizes the data.
 % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.

@@

 We start with some simple notation.  Let $N$ denote the population size.  At a given time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$. Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest. For simplicity, we start by ignoring the dynamic nature of the outbreak and recoverability of individuals. We assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$. We also let $I_j \in \{0,1\}$ be an indicator that the individual was tested ($I_j = 1$) or not ($I_j=0$).

 To start, we assume the overall number of active COVID-19 cases and/or active infection rate (AIR) are of primary interest. That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$. Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.  Then a natural candidate for AIR is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall active cases is $N \times \bar y$.
 Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.  Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{j=1}^N (Y_j - \bar Y)^2 = \bar Y (1- \bar Y)$.

@@

 Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and log odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}. Assuming no measurement error, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., if COVID-19 positive individuals were more likely to receive test then this implies upward bias in prevalence estimates. Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply magnifies this error. When tests are imperfect and selection bias exists, this simple relationship no longer holds.

@@

 \begin{figure}
 \centering
 \begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/sir_ratio.png}
  \caption{Ratio estimator}
  \label{fig:ratio-bias}
 \end{subfigure}%
 \begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/sir_rt.png}
  \caption{Effective reproductive rate estimator}
  \label{fig:r0-bias}
 \end{subfigure}
 \caption{Potential bias in ratio and effective reproductive rate estimators under an SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.  Here, $f = 0.02$, $FP = 0.024$, $FN = 0.13$, and a range of relative sampling fractions $M = f_1/f_0$ are considered.}
 \label{fig:rates}
 \end{figure}

@@

 % While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
 %Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
 %First, we can write the difference in the rates as $R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)$ where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
 %Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.  Here, these errors can mingle in interesting ways.
 Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.  Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model (A=black and B=grey). Figure~\ref{fig:eff} shows how biases interact in complex ways.  At first, the difference is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the magnitude of over-estimation increases as country A sees declining case-count while country B sees rapidly increasing case-count; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees declining case-count; finally, the gap disappears.  While this may not always be the case, the analysis demonstrates how estimates can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).

@@

 To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.  For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now. Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$. Here, we focus on logistic regression models, i.e., $\pi(x; \theta) = \text{expit} \left( x^\top \theta \right)$. Maximum likelihood estimation follows by maximizing
 \begin{equation}
 \label{eq:propensity}
 \sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
 \end{equation}
 The first sum only involves individuals observed in the nonprobability sample.
 The second sum is over the entire population.  Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.  Typically, this is not possible.  Here we present a method that uses auxiliary information obtained from probability samples.

@@

 As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
 % Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.  Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.
 Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer data size.

@@

 If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.  Unfortunately, large probabilistic samples are not available at every time within a given region.  To address this, here we consider a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
 $$
 \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
 $$
 where $K_h$ is a kernel function with tunable parameter $h$. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest} using case-count data observed on day~$t$.

 \subsubsection{Asymptotics}
 \label{section:asymptotics}

 Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.  For each finite population, we assume there are $L_\nu$ non-probability samples and probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^{L_{\nu}}$ over the study window~$[0,T]$. Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is consistent as $\nu$ goes to infinity (i.e., $N_{\nu}$ and $L_{\nu} \to \infty$) and calculates the estimator's variance.  In Lemma~\ref{lemma:ipw}, sensitivity and specificity are not assumed known.  Instead, we suppose that these are estimated using a pseudo-likelihood on two random samples from the population whose size (denoted~$n_{FP}$ and $n_{FN}$ respectively) is chosen so the resulting estimated sensitivity and specificity have desired levels of uncertainty.  See Appendix~\ref{app:asympderivations} for additional details.

 \begin{lemma}[Variance of IPW estimator] \normalfont
 \label{lemma:ipw}
 % \textcolor{red}{Must check.}
 The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ are solutions to the following set of estimating equations
 $$
 \Phi_n (\eta_t) =
 \left (
 \begin{array}{c}
 \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
 \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
 \frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
 \frac{1}{n_{FN}} \sum_{i=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
 \end{array}
 \right ) = {\bf 0}
 $$
 where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{h} (|t - t_l^\prime|)$ and $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
 \end{lemma}

@@

 Selection bias is addressed by analyzing COVID-19 death data rather than case counts.  Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$.  Then
 $$
 D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r \text{IFR}_{k} \cdot I_{k,t} \theta_{(r-t)} \right)
 $$
 where $\text{IFR}_{k}$ is the infection fatality rate for the $k$th strata and $\theta_{j}$ is a discrete-time distribution for time from infection to death.  This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on a stratification variable, which is important as IFR depends heavily on age~\citep{Levin2020}. See Section~\ref{section:application_of_modelbased} for discussion of parameter choices.  %Additional details on inference are discussed in Appendix~\ref{app:modelbased}.

@@

 In the first approach, we estimate two logistic regressions using weighted pseudo-likelihoods with the Facebook survey symptom data.  The first estimates the probability of contact with a COVID-19 positive individual given demographic and COVID-19 test status (positive or negative COVID-19 test). Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} in Section~\ref{app:in_add_details} in the supplementary materials presents the estimated likelihood of contact with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.  The second estimates the probability of fever given demographic, COVID-19 test status, \emph{and} whether the individual has had contact with a COVID-19 positive individual.  Figure~\ref{fig:symptomlik1} and~\ref{fig:symptomlik2} in Section~\ref{app:in_add_details} in the supplementary materials presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.  We see that likelihood of fever varies greatly based on whether the individual also reported contact with a COVID-19 positive individual. Using these two models, mean imputation is used to calculate the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status. Inverse probability weights are computed based on the resulting dataset (termed IPW1).
